---
title: "MTCL1 Cryptic Exon Analysis"
subtitle: "Characterization of cryptic exons 10a and 10b in NMD-inhibited lung cell lines"
author: "NMD Study"
date: "2026-03-01"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

<style>
body { font-size: 16px; }
.main-container { max-width: 1200px; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)
library(tidyverse)
library(DT)
library(knitr)
library(patchwork)
library(scales)
library(ggrepel)
library(Rsamtools)
library(GenomicAlignments)
library(Biostrings)

# BSgenome for genomic sequence extraction (PTC analysis)
if (!requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE))
  stop("Install with: BiocManager::install('BSgenome.Hsapiens.UCSC.hg38')")
library(BSgenome.Hsapiens.UCSC.hg38)

# maser for rMATS S4 object access
if (!requireNamespace("maser", quietly = TRUE))
  stop("Install with: BiocManager::install('maser')")
library(maser)
```

# Executive Summary

**MTCL1** (Microtubule Cross-Linking Factor 1; chr18, + strand) harbors two **cryptic exons**
(10a and 10b) in the intron between canonical exons 9 and 11. Analysis of PacBio long-read
BAM data from six lung cell types reveals that these exons are included more frequently under
**NMD inhibition** (Smg1i treatment), consistent with GENCODE's annotation of
ENST00000520495.5 (MTCL1-208) as a nonsense-mediated decay transcript.

Key findings:

- **Overall cryptic exon inclusion** increases from ~14% (DMSO) to ~18% (Smg1i) across cell types
- The strongest effects are in **DD** (11% → 27%) and **DD_ALI** (5% → 17%)
- Exon **10a** (114 bp) shows clean canonical splicing (exon 9 → 10a → exon 11)
- Exon **10b** (228 bp) is predominantly found in locally-initiating transcripts
- The Smg1i-driven increase is largely attributable to **local reads** starting near the cryptic region
- **PTC confirmation**: Genomic sequence analysis confirms premature termination codons in the cryptic exon inserts; the retained-intron structure (S3) introduces a frameshift guaranteeing a PTC, while in-frame structures (S1, S2) are evaluated for in-frame stop codons
- **PTC enrichment**: PTC-containing reads are enriched under Smg1i relative to non-PTC reads across spanning reads, directly supporting NMD as the degradation mechanism
- **Short-read corroboration**: rMATS confirms consistently positive dPSI for exon 10a across all 6 cell types, though none reach FDR significance individually

# Setup & Data Loading

## Constants and coordinates

```{r constants}
# Cryptic exon coordinates (chr18, + strand, 1-based inclusive)
EXON9_START  <- 8806893
EXON9_END    <- 8807060
EXON10A_START <- 8809448
EXON10A_END   <- 8809561
EXON10B_START <- 8809971
EXON10B_END   <- 8810198
EXON11_START  <- 8812979
EXON11_END    <- 8813233

# Junction tolerance for matching
JUNC_TOL <- 5

# Zoomed region for plots
ZOOM_START <- 8806000
ZOOM_END   <- 8814000

# TSS classification thresholds
TSS_FULLLENGTH_MAX <- 8720000
TSS_INTERNAL_MAX   <- 8806000

# Color palettes
treatment_colors <- c("DMSO" = "#2166AC", "Smg1i" = "#B2182B")
celltype_colors <- c(
  "AT" = "#E69F00", "DD" = "#56B4E9", "DD_ALI" = "#009E73",
  "DO" = "#F0E442", "FB" = "#0072B2", "MV" = "#D55E00"
)
exon_type_colors <- c("CDS" = "#2C3E50", "exon" = "#85929E")

# BAM → DGE cell type mapping
ct_to_dge <- c(
  "AT" = "at", "DD" = "dd", "DD_ALI" = "ddali",
  "DO" = "do", "FB" = "fb", "MV" = "mv"
)

# rMATS data source and cell type mapping
RMATS_RDS <- "../../results/rds/allrmats_jc_2025.8.20.rds"
rmats_ct_map <- c(AT2 = "AT", DD = "DD", DD_ALI = "DD_ALI",
                  DO_ALI = "DO", FB = "FB", MV = "MV")

# Coordinate summary table
exon_coords <- tibble(
  Exon = c("Exon 9 (canonical)", "Exon 10a (cryptic)", "Exon 10b (cryptic)", "Exon 11 (canonical)"),
  Start = c(EXON9_START, EXON10A_START, EXON10B_START, EXON11_START),
  End = c(EXON9_END, EXON10A_END, EXON10B_END, EXON11_END),
  Length_bp = End - Start + 1,
  `Divisible by 3` = (Length_bp %% 3 == 0)
)
```

## Load isoform annotation

```{r load-annotation}
ann <- read_tsv("isoform_annotation_MTCL1.tsv", show_col_types = FALSE)

# Parse exon coordinates from the exons column
parse_exons <- function(exon_str) {
  if (is.na(exon_str) || exon_str == "") return(tibble())
  entries <- str_split(exon_str, ";")[[1]]
  map_dfr(entries, function(e) {
    parts <- str_split(e, ":")[[1]]
    coords <- str_split(parts[1], "_")[[1]]
    tibble(
      start = as.numeric(coords[1]),
      end = as.numeric(coords[2]),
      strand = coords[3],
      type = parts[2]
    )
  })
}

# Parse all isoform exons
isoform_exons <- ann %>%
  select(isoform_id, isoform_name, source, exons) %>%
  mutate(parsed = map(exons, parse_exons)) %>%
  unnest(parsed) %>%
  select(-exons)
```

## Read BAM files

```{r load-bams}
# Search for BAMs in current dir and individual/ subdirectory (backward-compatible)
bam_search_dirs <- c(".", "individual", "merged")
all_bam_candidates <- unlist(lapply(bam_search_dirs, function(d) {
  if (dir.exists(d)) list.files(d, pattern = "\\.aligned\\.bam$", full.names = TRUE)
}))

# Prefer individual/ BAMs for per-sample data; fall back to flat layout
individual_bams <- grep("individual/", all_bam_candidates, value = TRUE)
flat_bams <- grep("^\\./[^/]+$", all_bam_candidates, value = TRUE)
bam_files <- if (length(individual_bams) > 0) individual_bams else flat_bams
# Exclude merged BAMs to avoid double-counting
bam_files <- bam_files[!grepl("merged/", bam_files)]

process_bam <- function(bam_path) {
  # Parse cell type, treatment, and sample_id from filename
  # Supports two formats:
  #   Merged:     AT_DMSO.aligned.bam       → cell_type="AT", treatment="DMSO", sample_id=NA
  #   Per-sample: Sample13_DD_017Q_DMSO.aligned.bam → cell_type="DD", treatment="DMSO",
  #               sample_id="Sample13_DD_017Q"
  base <- sub("\\.aligned\\.bam$", "", basename(bam_path))

  sample_id <- NA_character_
  donor <- NA_character_

  if (grepl("^Sample[0-9]+_", base)) {
    # Per-sample format: Sample{N}_{CT}_{DONOR}_{TREATMENT}
    m <- regmatches(base, regexec("^(Sample[0-9]+)_(.+)_(DMSO|Smg1i)$", base))[[1]]
    if (length(m) == 4) {
      treatment <- m[4]
      middle <- m[3]
      # middle is like "DD_017Q" or "DD_ALI_001V"
      donor <- sub(".*_", "", middle)
      cell_type <- sub(paste0("_", donor, "$"), "", middle)
      sample_id <- paste0(m[2], "_", middle)
    } else {
      # Fallback
      if (grepl("_Smg1i$", base)) {
        treatment <- "Smg1i"; cell_type <- sub("_Smg1i$", "", base)
      } else {
        treatment <- "DMSO"; cell_type <- sub("_DMSO$", "", base)
      }
    }
  } else {
    # Merged format: CT_TREATMENT
    if (grepl("_Smg1i$", base)) {
      treatment <- "Smg1i"
      cell_type <- sub("_Smg1i$", "", base)
    } else {
      treatment <- "DMSO"
      cell_type <- sub("_DMSO$", "", base)
    }
  }

  # Read alignments
  ga <- readGAlignments(bam_path, param = ScanBamParam(what = "qname"))
  n_reads <- length(ga)

  if (n_reads == 0) {
    return(tibble(
      qname = character(), cell_type = character(), treatment = character(),
      sample_id = character(), donor = character(),
      read_start = integer(), includes_10a = logical(), includes_10b = logical(),
      spans_e9_e11 = logical(), tss_category = character(),
      junctions = list()
    ))
  }

  # Define regions of interest
  e9_gr  <- GRanges("chr18", IRanges(EXON9_START, EXON9_END))
  e11_gr <- GRanges("chr18", IRanges(EXON11_START, EXON11_END))
  e10a_gr <- GRanges("chr18", IRanges(EXON10A_START, EXON10A_END))
  e10b_gr <- GRanges("chr18", IRanges(EXON10B_START, EXON10B_END))

  # Get aligned blocks for each read
  blocks_grl <- grglist(ga)

  # Check overlaps
  ov_e9  <- overlapsAny(blocks_grl, e9_gr)
  ov_e11 <- overlapsAny(blocks_grl, e11_gr)
  ov_10a <- overlapsAny(blocks_grl, e10a_gr)
  ov_10b <- overlapsAny(blocks_grl, e10b_gr)

  # Extract junctions from each read
  get_read_junctions <- function(blocks) {
    blocks <- sort(blocks)
    if (length(blocks) < 2) return(tibble(donor = integer(), acceptor = integer()))
    tibble(
      donor = end(blocks)[-length(blocks)],
      acceptor = start(blocks)[-1]
    )
  }

  junc_list <- lapply(blocks_grl, get_read_junctions)

  # Classify TSS
  read_starts <- start(ga)
  tss_cat <- case_when(
    read_starts < TSS_FULLLENGTH_MAX ~ "full-length",
    read_starts < TSS_INTERNAL_MAX   ~ "internal",
    TRUE                              ~ "local"
  )

  tibble(
    qname = mcols(ga)$qname,
    cell_type = cell_type,
    treatment = treatment,
    sample_id = sample_id,
    donor = donor,
    read_start = read_starts,
    includes_10a = ov_10a,
    includes_10b = ov_10b,
    spans_e9_e11 = ov_e9 & ov_e11,
    tss_category = tss_cat,
    junctions = junc_list
  )
}

# Process all BAMs
all_reads <- map_dfr(bam_files, process_bam)

# Read count summary
read_counts <- all_reads %>%
  dplyr::count(cell_type, treatment, name = "total_reads") %>%
  arrange(cell_type, treatment)
```

## Load DGE results

```{r load-dge}
dge_files <- list.files("../../isocall_dge", pattern = "nmd_isocall_dge_.*_2026\\.3\\.1\\.csv",
                        full.names = TRUE)

load_dge <- function(path) {
  ct_code <- str_extract(basename(path), "(?<=nmd_isocall_dge_)[a-z0-9]+(?=_)")
  ct_label <- names(ct_to_dge)[ct_to_dge == ct_code]
  if (length(ct_label) != 1) return(tibble())
  df <- read_csv(path, show_col_types = FALSE)
  df %>%
    filter(gene_name == "MTCL1") %>%
    select(any_of(c("transcript_id", "gene_name", "logFC", "CI.L", "CI.R",
                     "AveExpr", "P.Value", "adj.P.Val", "B"))) %>%
    mutate(cell_type = ct_label)
}

dge_mtcl1 <- map_dfr(dge_files, load_dge)
```

## Read count summary

```{r read-count-table}
read_counts %>%
  pivot_wider(names_from = treatment, values_from = total_reads) %>%
  mutate(Total = DMSO + Smg1i) %>%
  kable(caption = "Total reads per BAM file")
```

# MTCL1 Gene & Isoform Overview

## Isoform models

```{r fig1-gene-structure, fig.height=8, fig.cap="Figure 1: MTCL1 gene structure showing top expressed isoforms plus NMD-annotated transcripts. Cryptic exon region (exons 10a/10b) highlighted in red."}
# Select isoforms to display: top expressed + key NMD isoforms
top_expressed <- ann %>%
  filter(total_reads_all_samples > 0) %>%
  arrange(desc(total_reads_all_samples)) %>%
  head(8)

# Ensure MTCL1-208 and MTCL1-206 are included
key_isoforms <- c("ENST00000520495.5", "ENST00000518815.1")
display_ids <- union(top_expressed$isoform_id, key_isoforms)

display_ann <- ann %>% filter(isoform_id %in% display_ids)

# Order by TSS position (descending for plot)
display_ann <- display_ann %>%
  arrange(desc(tss)) %>%
  mutate(
    y_pos = row_number(),
    label = case_when(
      isoform_id == "ENST00000520495.5" ~ paste0(isoform_name, " (NMD)"),
      isoform_id == "ENST00000518815.1" ~ paste0(isoform_name, " (10a)"),
      TRUE ~ isoform_name
    )
  )

# Get exons for display isoforms
display_exons <- isoform_exons %>%
  filter(isoform_id %in% display_ids) %>%
  left_join(display_ann %>% select(isoform_id, y_pos, label), by = "isoform_id")

# Gene extent
gene_start <- min(ann$tss, na.rm = TRUE)
gene_end <- max(ann$tes, na.rm = TRUE)

ggplot() +
  # Cryptic region highlight
  annotate("rect",
    xmin = EXON10A_START - 500, xmax = EXON10B_END + 500,
    ymin = 0.3, ymax = max(display_ann$y_pos) + 0.7,
    fill = "#B2182B", alpha = 0.08
  ) +
  # Intron lines
  geom_segment(
    data = display_ann,
    aes(x = tss, xend = tes, y = y_pos, yend = y_pos),
    color = "grey70", linewidth = 0.3
  ) +
  # Exon rectangles
  geom_rect(
    data = display_exons,
    aes(xmin = start, xmax = end, ymin = y_pos - 0.3, ymax = y_pos + 0.3,
        fill = type),
    color = "grey30", linewidth = 0.2
  ) +
  scale_fill_manual(values = exon_type_colors, name = "Exon type") +
  scale_y_continuous(
    breaks = display_ann$y_pos,
    labels = display_ann$label
  ) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "MTCL1 gene structure",
    subtitle = "Top expressed isoforms + NMD-annotated transcripts",
    x = "chr18 position", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10)
  )
```

# Cryptic Exon Analysis

## Genomic context

```{r exon-coord-table}
exon_coords %>%
  mutate(Start = format(Start, big.mark = ","), End = format(End, big.mark = ",")) %>%
  kable(caption = "Cryptic exon coordinates (chr18, GRCh38)")
```

Two GENCODE isoforms use portions of this intronic region, but neither matches the
structures observed in the raw reads:

- **MTCL1-208** (ENST00000520495.5) is annotated as `nonsense_mediated_decay`.
  It includes a single terminal exon at 8,809,448--8,810,002 (555 bp) that overlaps 10a and
  part of 10b, but the transcript **ends there** --- it never splices to exon 11.
- **MTCL1-206** (ENST00000518815.1) includes exon 10a (8,809,448--8,809,561) and
  **does** splice it to exon 11, matching one of the read-supported structures (see Section 5).
  However, it has no 10b.

Exon 10b (8,809,971--8,810,198) extends 196 bp past the end of MTCL1-208's terminal exon,
meaning it is not fully contained in any GENCODE transcript model. The discrete two-cassette
architecture (exon 9 → 10a → 10b → exon 11) observed in multiple Smg1i-enriched reads
is likewise absent from GENCODE.

Both exon 10a (114 bp) and exon 10b (228 bp) are divisible by 3, so their inclusion
maintains the reading frame if canonical splice sites are used.

```{r fig3-zoomed-structure, fig.height=3.5, fig.cap="Figure 3: Zoomed view of the intron between exons 9 and 11, showing cryptic exons 10a and 10b."}
exon_rects <- tibble(
  label = c("Exon 9", "Exon 10a", "Exon 10b", "Exon 11"),
  start = c(EXON9_START, EXON10A_START, EXON10B_START, EXON11_START),
  end = c(EXON9_END, EXON10A_END, EXON10B_END, EXON11_END),
  fill = c("canonical", "cryptic", "cryptic", "canonical"),
  y = 0
)

ggplot(exon_rects) +
  # Connecting line (intron)
  annotate("segment", x = EXON9_END, xend = EXON11_START, y = 0, yend = 0,
           color = "grey60", linewidth = 0.5) +
  # Exon rectangles
  geom_rect(aes(xmin = start, xmax = end, ymin = -0.3, ymax = 0.3, fill = fill),
            color = "grey30", linewidth = 0.4) +
  # Labels
 geom_text(aes(x = (start + end) / 2, y = 0.55, label = label), size = 3.5, fontface = "bold") +
  # Size annotations
  geom_text(
    aes(x = (start + end) / 2, y = -0.55,
        label = paste0(end - start + 1, " bp")),
    size = 3, color = "grey40"
  ) +
  scale_fill_manual(
    values = c("canonical" = "#2C3E50", "cryptic" = "#E74C3C"),
    guide = "none"
  ) +
  scale_x_continuous(
    limits = c(ZOOM_START, ZOOM_END),
    labels = label_comma()
  ) +
  labs(
    title = "MTCL1 intron 9–11: cryptic exon region",
    subtitle = "chr18:8,806,000-8,814,000 (GRCh38)",
    x = "chr18 position", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(ylim = c(-1, 1))
```

## Cryptic exon inclusion rates

```{r inclusion-analysis}
# Denominator: reads spanning exon 9 and exon 11
spanning <- all_reads %>% filter(spans_e9_e11)

# Classify cryptic exon inclusion
spanning <- spanning %>%
  mutate(
    cryptic_category = case_when(
      includes_10a & includes_10b ~ "both",
      includes_10a ~ "10a only",
      includes_10b ~ "10b only",
      TRUE ~ "none"
    ),
    has_cryptic = cryptic_category != "none"
  )

# Summary per cell type and treatment
inclusion_summary <- spanning %>%
  group_by(cell_type, treatment) %>%
  summarise(
    n_spanning = n(),
    n_cryptic = sum(has_cryptic),
    n_10a_only = sum(cryptic_category == "10a only"),
    n_10b_only = sum(cryptic_category == "10b only"),
    n_both = sum(cryptic_category == "both"),
    pct_cryptic = 100 * n_cryptic / n_spanning,
    .groups = "drop"
  )

# Overall summary
overall_summary <- spanning %>%
  group_by(treatment) %>%
  summarise(
    n_spanning = n(),
    n_cryptic = sum(has_cryptic),
    pct_cryptic = 100 * n_cryptic / n_spanning,
    .groups = "drop"
  )
```

```{r fig4-inclusion-barplot, fig.height=5, fig.cap="Figure 4: Cryptic exon inclusion rates across cell types. Denominator = reads spanning both exon 9 and exon 11."}
ggplot(inclusion_summary, aes(x = cell_type, y = pct_cryptic, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = paste0(round(pct_cryptic, 1), "%\n(", n_cryptic, "/", n_spanning, ")")),
    position = position_dodge(width = 0.7),
    vjust = -0.3, size = 3
  ) +
  scale_fill_manual(values = treatment_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Cryptic exon inclusion rate by cell type",
    subtitle = "% of exon 9–11 spanning reads that include exon 10a and/or 10b",
    x = "Cell type", y = "Inclusion rate (%)", fill = "Treatment"
  ) +
  theme_minimal(base_size = 13)
```

```{r fisher-tests}
# Fisher's exact test for each cell type
fisher_results <- inclusion_summary %>%
  select(cell_type, treatment, n_cryptic, n_spanning) %>%
  mutate(n_no_cryptic = n_spanning - n_cryptic) %>%
  pivot_wider(
    names_from = treatment,
    values_from = c(n_cryptic, n_no_cryptic, n_spanning)
  ) %>%
  rowwise() %>%
  mutate(
    fisher_p = fisher.test(matrix(
      c(n_cryptic_DMSO, n_no_cryptic_DMSO, n_cryptic_Smg1i, n_no_cryptic_Smg1i),
      nrow = 2
    ))$p.value
  ) %>%
  ungroup()

# Wilson CIs for each group
wilson_ci <- inclusion_summary %>%
  rowwise() %>%
  mutate(
    ci = list(prop.test(n_cryptic, n_spanning, correct = FALSE)),
    ci_low = ci$conf.int[1] * 100,
    ci_high = ci$conf.int[2] * 100
  ) %>%
  ungroup() %>%
  select(cell_type, treatment, pct_cryptic, ci_low, ci_high)

# Combined table
inclusion_table <- inclusion_summary %>%
  left_join(wilson_ci, by = c("cell_type", "treatment", "pct_cryptic")) %>%
  mutate(
    rate = sprintf("%.1f%% [%.1f–%.1f]", pct_cryptic, ci_low, ci_high)
  ) %>%
  select(cell_type, treatment, n_spanning, n_cryptic, n_10a_only, n_10b_only, n_both, rate) %>%
  left_join(
    fisher_results %>% select(cell_type, fisher_p),
    by = "cell_type"
  )

inclusion_table %>%
  mutate(fisher_p = ifelse(treatment == "DMSO", signif(fisher_p, 3), "")) %>%
  kable(
    caption = "Cryptic exon inclusion rates with 95% Wilson CIs and Fisher's exact test p-values",
    col.names = c("Cell type", "Treatment", "Spanning", "Cryptic", "10a only", "10b only",
                  "Both", "Rate [95% CI]", "Fisher p")
  )
```

## Splice junctions

```{r junction-analysis}
# Extract all junctions from spanning reads
spanning_juncs <- spanning %>%
  select(qname, cell_type, treatment, junctions, cryptic_category) %>%
  unnest(junctions)

# Define key junctions
key_junctions <- tribble(
  ~junction_name, ~target_donor, ~target_acceptor,
  "Canonical skip (9→11)", EXON9_END, EXON11_START,
  "Into 10a (9→10a)", EXON9_END, EXON10A_START,
  "Out of 10a (10a→11)", EXON10A_END, EXON11_START,
  "10a→10b", EXON10A_END, EXON10B_START,
  "Into 10b (9→10b)", EXON9_END, EXON10B_START,
  "Out of 10b (10b→11)", EXON10B_END, EXON11_START
)

# Match junctions with tolerance
classify_junction <- function(donor, acceptor) {
  matches <- key_junctions %>%
    filter(abs(target_donor - donor) <= JUNC_TOL & abs(target_acceptor - acceptor) <= JUNC_TOL)
  if (nrow(matches) > 0) matches$junction_name[1] else "other"
}

junction_counts <- spanning_juncs %>%
  # Filter to junctions in the region of interest
  filter(donor >= EXON9_START - 100, acceptor <= EXON11_END + 100) %>%
  rowwise() %>%
  mutate(junction_name = classify_junction(donor, acceptor)) %>%
  ungroup() %>%
  dplyr::count(cell_type, treatment, junction_name) %>%
  arrange(cell_type, treatment, desc(n))
```

```{r fig5-sashimi, fig.height=6, fig.cap="Figure 5: Sashimi-style junction arc plot for DD cells. Arc thickness proportional to junction read count. Exon rectangles shown at bottom."}
# Focus on DD for sashimi plot
dd_juncs <- spanning_juncs %>%
  filter(cell_type == "DD") %>%
  filter(donor >= EXON9_START - 100, acceptor <= EXON11_END + 100) %>%
  rowwise() %>%
  mutate(junction_name = classify_junction(donor, acceptor)) %>%
  ungroup() %>%
  filter(junction_name != "other") %>%
  dplyr::count(treatment, junction_name, donor, acceptor) %>%
  # Use median positions for matching junctions
  group_by(treatment, junction_name) %>%
  summarise(
    donor = round(median(donor)),
    acceptor = round(median(acceptor)),
    n = sum(n),
    .groups = "drop"
  )

# Generate arc coordinates
make_arc <- function(x_start, x_end, count, n_points = 100) {
  t <- seq(0, pi, length.out = n_points)
  tibble(
    x = x_start + (x_end - x_start) * (t / pi),
    y = sin(t) * sqrt(count) * 0.5
  )
}

arc_data <- dd_juncs %>%
  mutate(arc_id = row_number()) %>%
  rowwise() %>%
  mutate(arc = list(make_arc(donor, acceptor, n))) %>%
  unnest(arc)

# Exon rectangles for the sashimi plot
sashimi_exons <- tibble(
  label = c("Exon 9", "10a", "10b", "Exon 11"),
  start = c(EXON9_START, EXON10A_START, EXON10B_START, EXON11_START),
  end = c(EXON9_END, EXON10A_END, EXON10B_END, EXON11_END),
  fill = c("canonical", "cryptic", "cryptic", "canonical")
)

# Junction labels
junc_labels <- dd_juncs %>%
  mutate(
    x_mid = (donor + acceptor) / 2,
    y_top = sqrt(n) * 0.5 + 0.15,
    label = paste0(junction_name, "\n(n=", n, ")")
  )

ggplot() +
  # Exon rectangles at bottom
  geom_rect(
    data = sashimi_exons,
    aes(xmin = start, xmax = end, ymin = -0.4, ymax = -0.1, fill = fill),
    color = "grey30", linewidth = 0.3
  ) +
  # Intron line
  annotate("segment", x = EXON9_END, xend = EXON11_START,
           y = -0.25, yend = -0.25, color = "grey60", linewidth = 0.3) +
  # Junction arcs
  geom_path(
    data = arc_data,
    aes(x = x, y = y, group = interaction(treatment, arc_id),
        color = treatment, linewidth = n)
  ) +
  # Junction count labels
  geom_text_repel(
    data = junc_labels,
    aes(x = x_mid, y = y_top, label = label, color = treatment),
    size = 2.8, show.legend = FALSE, direction = "y",
    max.overlaps = 20
  ) +
  facet_wrap(~treatment, ncol = 1) +
  scale_fill_manual(values = c("canonical" = "#2C3E50", "cryptic" = "#E74C3C"), guide = "none") +
  scale_color_manual(values = treatment_colors, guide = "none") +
  scale_linewidth_continuous(range = c(0.5, 3), guide = "none") +
  scale_x_continuous(
    limits = c(ZOOM_START, ZOOM_END),
    labels = label_comma()
  ) +
  labs(
    title = "Splice junction usage in DD cells",
    subtitle = "Arc height ∝ √(count); zoomed to exon 9–11 region",
    x = "chr18 position", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r junction-table}
# Junction count table across all cell types
junction_counts %>%
  pivot_wider(names_from = treatment, values_from = n, values_fill = 0) %>%
  arrange(cell_type, desc(DMSO + Smg1i)) %>%
  kable(caption = "Splice junction counts in the exon 9–11 region")
```

## TSS classification

```{r tss-analysis}
# TSS categories for ALL reads with cryptic exon overlap (not just spanning)
# Local reads often start near the cryptic region and may not extend to exon 11,
# so requiring spanning would undercount them
cryptic_all <- all_reads %>% filter(includes_10a | includes_10b)

tss_summary <- cryptic_all %>%
  dplyr::count(treatment, tss_category) %>%
  complete(treatment, tss_category = c("full-length", "internal", "local"), fill = list(n = 0))

tss_totals <- tss_summary %>%
  group_by(treatment) %>%
  summarise(total = sum(n), .groups = "drop")
```

```{r fig6-tss-barplot, fig.height=5, fig.cap="Figure 6: TSS classification of cryptic-exon-including reads. The Smg1i increase is predominantly driven by local reads initiating near the cryptic exon region."}
tss_colors <- c("full-length" = "#2C3E50", "internal" = "#3498DB", "local" = "#E74C3C")

ggplot(tss_summary, aes(x = treatment, y = n, fill = tss_category)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = ifelse(n > 0, n, "")),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 4
  ) +
  geom_text(
    data = tss_totals,
    aes(x = treatment, y = total, label = paste0("n=", total), fill = NULL),
    vjust = -0.5, size = 4
  ) +
  scale_fill_manual(
    values = tss_colors,
    name = "TSS category",
    labels = c(
      "full-length" = paste0("Full-length (TSS < ", format(TSS_FULLLENGTH_MAX, big.mark = ","), ")"),
      "internal" = "Internal TSS",
      "local" = paste0("Local (TSS ≥ ", format(TSS_INTERNAL_MAX, big.mark = ","), ")")
    )
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "TSS classification of cryptic-exon-including reads",
    subtitle = "Across all cell types",
    x = "Treatment", y = "Number of reads"
  ) +
  theme_minimal(base_size = 13)
```

# Short-Read Cross-Validation

## Short-read differential splicing (rMATS)

rMATS quantifies differential splicing from short-read RNA-seq using junction read counts.
Here we extract MTCL1 skipped-exon (SE) events matching exon 10a and compare dPSI values
with long-read inclusion rates. Positive dPSI indicates more exon inclusion under Smg1i
(condition 1 = Smg1i, condition 2 = DMSO).

```{r load-rmats}
# Load rMATS maser S4 objects (one per cell type)
rmats_available <- file.exists(RMATS_RDS)

if (rmats_available) {
  rmats_jc <- readRDS(RMATS_RDS)

  # Extract MTCL1 SE events and per-sample PSI across all cell types
  rmats_results <- list()
  rmats_psi_results <- list()

  for (ct_rmats in names(rmats_jc)) {
    obj <- rmats_jc[[ct_rmats]]
    ct <- rmats_ct_map[ct_rmats]

    events <- slot(obj, "SE_events") %>% as_tibble()
    stats  <- slot(obj, "SE_stats")  %>% as_tibble()
    psi    <- slot(obj, "SE_PSI")
    gr     <- slot(obj, "SE_gr")

    # Find MTCL1 SE events
    gene_idx <- which(events$geneSymbol == "MTCL1")
    if (length(gene_idx) == 0) next

    # Match exon 10a coordinates to the exon_target GRanges
    exon_target <- gr[["exon_target"]][gene_idx]
    target_starts <- start(exon_target)
    target_ends   <- end(exon_target)

    match_10a <- which(
      abs(target_starts - EXON10A_START) <= JUNC_TOL &
      abs(target_ends   - EXON10A_END)   <= JUNC_TOL
    )

    if (length(match_10a) == 0) next

    # Use the first matching event
    idx <- gene_idx[match_10a[1]]
    merged <- inner_join(
      events[idx, , drop = FALSE],
      stats[idx, , drop = FALSE],
      by = "ID"
    )

    rmats_results[[ct]] <- tibble(
      cell_type = ct,
      event_id  = merged$ID,
      dPSI      = merged$IncLevelDifference,
      FDR       = merged$FDR,
      PValue    = merged$PValue
    )

    # Extract per-sample PSI values
    n1 <- slot(obj, "n_cond1")
    n2 <- slot(obj, "n_cond2")
    psi_vals <- as.numeric(psi[idx, ])
    smg1i_psi <- psi_vals[1:n1]
    dmso_psi  <- psi_vals[(n1 + 1):(n1 + n2)]

    rmats_psi_results[[paste0(ct, "_Smg1i")]] <- tibble(
      cell_type = ct, treatment = "Smg1i",
      PSI = smg1i_psi[!is.na(smg1i_psi)],
      dPSI = merged$IncLevelDifference, FDR = merged$FDR
    )
    rmats_psi_results[[paste0(ct, "_DMSO")]] <- tibble(
      cell_type = ct, treatment = "DMSO",
      PSI = dmso_psi[!is.na(dmso_psi)],
      dPSI = merged$IncLevelDifference, FDR = merged$FDR
    )
  }

  rmats_summary <- bind_rows(rmats_results)
  rmats_psi <- bind_rows(rmats_psi_results)
} else {
  warning("rMATS RDS not found at: ", RMATS_RDS)
  rmats_summary <- tibble(
    cell_type = character(), event_id = integer(),
    dPSI = numeric(), FDR = numeric(), PValue = numeric()
  )
  rmats_psi <- tibble(
    cell_type = character(), treatment = character(),
    PSI = numeric(), dPSI = numeric(), FDR = numeric()
  )
}
```

```{r rmats-comparison-table, eval=nrow(rmats_summary) > 0}
# Join rMATS summary with long-read inclusion rates
lr_wide <- inclusion_summary %>%
  select(cell_type, treatment, pct_cryptic) %>%
  pivot_wider(names_from = treatment, values_from = pct_cryptic,
              names_prefix = "lr_pct_")

comparison_table <- rmats_summary %>%
  left_join(lr_wide, by = "cell_type") %>%
  mutate(
    lr_delta = lr_pct_Smg1i - lr_pct_DMSO,
    `Cell Type` = cell_type,
    `rMATS dPSI` = sprintf("%+.3f", dPSI),
    `rMATS FDR` = formatC(FDR, format = "e", digits = 1),
    `LR incl. DMSO` = sprintf("%.1f%%", lr_pct_DMSO),
    `LR incl. Smg1i` = sprintf("%.1f%%", lr_pct_Smg1i),
    `LR Δ` = sprintf("%+.1f%%", lr_delta)
  ) %>%
  select(`Cell Type`, `rMATS dPSI`, `rMATS FDR`,
         `LR incl. DMSO`, `LR incl. Smg1i`, `LR Δ`)

kable(comparison_table, caption = "Cross-platform comparison: rMATS short-read dPSI vs long-read cryptic exon inclusion rates for exon 10a")
```

```{r fig-rmats-comparison, fig.width=12, fig.height=5, eval=nrow(rmats_psi) > 0}
# Panel A: rMATS per-replicate PSI
rmats_means <- rmats_psi %>%
  group_by(cell_type, treatment) %>%
  summarise(mean_PSI = mean(PSI), .groups = "drop")

p_rmats <- ggplot(rmats_psi, aes(x = treatment, y = PSI, color = treatment)) +
  geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.7) +
  geom_crossbar(data = rmats_means,
                aes(x = treatment, y = mean_PSI, ymin = mean_PSI, ymax = mean_PSI),
                width = 0.4, linewidth = 0.6) +
  facet_wrap(~cell_type, nrow = 1) +
  scale_color_manual(values = treatment_colors) +
  scale_y_continuous(labels = scales::percent_format(scale = 100),
                     limits = c(0, NA)) +
  labs(title = "A) Short-read rMATS PSI (exon 10a)",
       subtitle = "Per-replicate PSI values with mean crossbar",
       x = "Treatment", y = "PSI (Percent Spliced In)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))

# Panel B: Long-read per-sample inclusion rate
# Use per-sample data if sample_id is available, otherwise fall back to group-level
has_per_sample <- "sample_id" %in% names(all_reads) && any(!is.na(all_reads$sample_id))

if (has_per_sample) {
  lr_per_sample <- all_reads %>%
    filter(spans_e9_e11) %>%
    mutate(has_cryptic = includes_10a | includes_10b) %>%
    group_by(cell_type, treatment, sample_id) %>%
    summarise(
      n_spanning = n(),
      n_cryptic = sum(has_cryptic),
      pct_cryptic = 100 * n_cryptic / n_spanning,
      .groups = "drop"
    )

  lr_means <- lr_per_sample %>%
    group_by(cell_type, treatment) %>%
    summarise(mean_pct = mean(pct_cryptic), .groups = "drop")

  p_lr <- ggplot(lr_per_sample, aes(x = treatment, y = pct_cryptic, color = treatment)) +
    geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.7) +
    geom_crossbar(data = lr_means,
                  aes(x = treatment, y = mean_pct, ymin = mean_pct, ymax = mean_pct),
                  width = 0.4, linewidth = 0.6) +
    facet_wrap(~cell_type, nrow = 1) +
    scale_color_manual(values = treatment_colors) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = "B) Long-read cryptic exon inclusion rate",
         subtitle = "Per-sample inclusion rates with mean crossbar",
         x = "Treatment", y = "Inclusion rate (%)") +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none",
          strip.text = element_text(face = "bold"))
} else {
  # Fall back to group-level bars if no per-sample data
  p_lr <- ggplot(inclusion_summary, aes(x = treatment, y = pct_cryptic, fill = treatment)) +
    geom_col(width = 0.6) +
    facet_wrap(~cell_type, nrow = 1) +
    scale_fill_manual(values = treatment_colors) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = "B) Long-read cryptic exon inclusion rate",
         subtitle = "Group-level inclusion rates (merged BAMs)",
         x = "Treatment", y = "Inclusion rate (%)") +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none",
          strip.text = element_text(face = "bold"))
}

p_rmats + p_lr + plot_layout(widths = c(1, 1))
```

```{r rmats-discussion, results='asis', eval=nrow(rmats_summary) > 0}
n_positive <- sum(rmats_summary$dPSI > 0)
n_total <- nrow(rmats_summary)
n_sig <- sum(rmats_summary$FDR < 0.05)
mean_dpsi <- mean(rmats_summary$dPSI)

cat(sprintf(
"### Cross-platform concordance

rMATS identifies a skipped-exon event matching exon 10a (chr18:%s–%s) in **%d of %d**
cell types with **consistently positive dPSI** (mean = %+.3f), indicating increased
inclusion under Smg1i treatment. %s

This concordance between short-read junction counts (rMATS) and long-read full-length
transcript analysis strengthens the evidence for NMD-mediated regulation of cryptic
exon 10a inclusion.
",
format(EXON10A_START, big.mark = ","), format(EXON10A_END, big.mark = ","),
n_positive, n_total, mean_dpsi,
if (n_sig > 0) {
  sprintf("**%d** cell type(s) reach FDR < 0.05.", n_sig)
} else {
  "However, none reach FDR < 0.05 individually, likely due to modest effect sizes and limited replicates in the short-read experiment."
}
))

cat("\n\n**Limitations of short-read comparison:**\n\n")
cat("- rMATS can only assess **exon 10a** as a simple skipped exon; exon 10b and multi-exon ")
cat("structures (S2, S3, S4) are not detectable from junction counts alone.\n")
cat("- Short-read PSI is junction-count-based, while long-read inclusion is read-spanning-based; ")
cat("these denominators are not directly comparable.\n")
cat("- The reading frame classification (PTC status) confirmed in the long-read analysis ")
cat("cannot be determined from short-read splicing data alone.\n")
```

## Short-read junction read analysis {.tabset}

This section is conditional on the availability of short-read junction BAMs
(produced by `extract_shortread_junctions.sh`).

```{r junction-gate}
has_junction_bams <- length(
  list.files(".", pattern = "\\.junction\\.bam$", recursive = TRUE)
) > 0
```

```{r process-junction-bams, eval=has_junction_bams}
# Discover junction BAMs (individual/ subdirectory preferred)
jbam_search_dirs <- c(".", "individual")
jbam_files <- unlist(lapply(jbam_search_dirs, function(d) {
  if (dir.exists(d)) list.files(d, pattern = "\\.junction\\.bam$", full.names = TRUE)
}))
jbam_files <- jbam_files[!grepl("/merged/", jbam_files)]

process_junction_bam <- function(bam_path) {
  # Parse sample name from filename (e.g., Sample1_DD_017Q_DMSO.junction.bam)
  base <- sub("\\.junction\\.bam$", "", basename(bam_path))

  sample_id <- NA_character_
  donor <- NA_character_
  cell_type <- NA_character_
  treatment <- NA_character_

  if (grepl("^Sample[0-9]+_", base)) {
    m <- regmatches(base, regexec("^(Sample[0-9]+)_(.+)_(DMSO|Smg1i)$", base))[[1]]
    if (length(m) == 4) {
      treatment <- m[4]
      middle <- m[3]
      donor <- sub(".*_", "", middle)
      cell_type <- sub(paste0("_", donor, "$"), "", middle)
      sample_id <- paste0(m[2], "_", middle)
    }
  } else {
    if (grepl("_Smg1i$", base)) {
      treatment <- "Smg1i"; cell_type <- sub("_Smg1i$", "", base)
    } else {
      treatment <- "DMSO"; cell_type <- sub("_DMSO$", "", base)
    }
  }

  # Read paired-end alignments to avoid double-counting mates
  ga <- readGAlignmentPairs(bam_path, param = ScanBamParam(what = "qname"))
  n_pairs <- length(ga)

  if (n_pairs == 0) {
    return(tibble(
      cell_type = character(), treatment = character(), sample_id = character(),
      junction_type = character(), n_reads = integer()
    ))
  }

  # Extract junctions from first-in-pair for counting
  first_ga <- GenomicAlignments::first(ga)
  blocks_grl <- grglist(first_ga)

  get_junctions <- function(blocks) {
    blocks <- sort(blocks)
    if (length(blocks) < 2) return(tibble(donor = integer(), acceptor = integer()))
    tibble(
      donor = end(blocks)[-length(blocks)],
      acceptor = start(blocks)[-1]
    )
  }

  all_juncs <- bind_rows(lapply(blocks_grl, get_junctions))

  # Classify junctions relative to MTCL1 cryptic exons
  # Canonical skip: exon 9 → exon 11
  n_canonical_skip <- sum(
    abs(all_juncs$donor - EXON9_END) <= JUNC_TOL &
    abs(all_juncs$acceptor - EXON11_START) <= JUNC_TOL
  )

  # 10a inclusion: exon 9 → 10a
  n_e9_10a <- sum(
    abs(all_juncs$donor - EXON9_END) <= JUNC_TOL &
    abs(all_juncs$acceptor - EXON10A_START) <= JUNC_TOL
  )

  # 10a → exon 11
  n_10a_e11 <- sum(
    abs(all_juncs$donor - EXON10A_END) <= JUNC_TOL &
    abs(all_juncs$acceptor - EXON11_START) <= JUNC_TOL
  )

  tibble(
    cell_type = cell_type,
    treatment = treatment,
    sample_id = sample_id,
    junction_type = c("canonical_skip", "e9_to_10a", "e10a_to_e11"),
    n_reads = c(n_canonical_skip, n_e9_10a, n_10a_e11)
  )
}

sr_junctions <- map_dfr(jbam_files, process_junction_bam)

# Compute inclusion ratio per sample: inclusion / (inclusion + skip)
sr_inclusion <- sr_junctions %>%
  pivot_wider(names_from = junction_type, values_from = n_reads, values_fill = 0) %>%
  mutate(
    inclusion_junctions = pmin(e9_to_10a, e10a_to_e11),
    total_junctions = inclusion_junctions + canonical_skip,
    inclusion_ratio = ifelse(total_junctions > 0,
                             100 * inclusion_junctions / total_junctions, NA_real_)
  )
```

```{r fig-three-way-comparison, fig.width=14, fig.height=5, eval=has_junction_bams && nrow(rmats_psi) > 0}
# Three-way comparison: rMATS dPSI, SR junction ratio, LR inclusion rate
p1 <- ggplot(rmats_psi, aes(x = treatment, y = PSI * 100, color = treatment)) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.6) +
  facet_wrap(~cell_type, nrow = 1) +
  scale_color_manual(values = treatment_colors) +
  labs(title = "A) rMATS PSI", y = "PSI (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

p2 <- ggplot(sr_inclusion %>% filter(!is.na(inclusion_ratio)),
             aes(x = treatment, y = inclusion_ratio, color = treatment)) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.6) +
  facet_wrap(~cell_type, nrow = 1) +
  scale_color_manual(values = treatment_colors) +
  labs(title = "B) SR junction ratio", y = "Inclusion (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

if (has_per_sample) {
  p3 <- ggplot(lr_per_sample, aes(x = treatment, y = pct_cryptic, color = treatment)) +
    geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
    stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.6) +
    facet_wrap(~cell_type, nrow = 1) +
    scale_color_manual(values = treatment_colors) +
    labs(title = "C) LR inclusion rate", y = "Inclusion (%)") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none", strip.text = element_text(face = "bold"))
} else {
  p3 <- ggplot(inclusion_summary, aes(x = treatment, y = pct_cryptic, fill = treatment)) +
    geom_col(width = 0.6) +
    facet_wrap(~cell_type, nrow = 1) +
    scale_fill_manual(values = treatment_colors) +
    labs(title = "C) LR inclusion rate", y = "Inclusion (%)") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none", strip.text = element_text(face = "bold"))
}

p1 + p2 + p3 + plot_layout(widths = c(1, 1, 1))
```

# Biological Interpretation

## Raw reads vs. transcript models

The splicing patterns observed in the raw BAM data differ in important ways from existing
transcript annotations. Neither the GENCODE reference models nor the PacBio-called isoforms
(which are under active revision) fully capture the structures present in the reads.

**MTCL1-208** (the only GENCODE NMD-annotated isoform in this region) has a single 555 bp
terminal exon (8,809,448--8,810,002) that overlaps 10a and part of 10b. Crucially, this
transcript ends at 8,810,002 and never splices to exon 11. The raw reads, by contrast, show
cryptic exon inclusion as **internal cassette events** within longer transcripts that continue
through exon 11 and beyond. **MTCL1-206** does splice exon 10a to exon 11, matching one
read-supported pattern, but it is annotated as `protein_coding_CDS_not_defined` and contains no
10b. Exon 10b itself (ending at 8,810,198) extends past MTCL1-208's terminal exon boundary,
placing it outside any annotated transcript model.

In short, the transcript models provide partial precedent for cryptic exon usage in this
region, but the read data reveals structures --- particularly the discrete 10a + 10b
cassette --- that are not represented in current annotations.

## Read-supported isoform structures

Rather than relying on existing transcript models, we classify each cryptic-exon-including
read into one of four structural categories based on its aligned blocks and splice junctions
in the exon 9--11 region.

```{r structure-classification}
# Classify each cryptic-region read into a structural category
# We need all reads that overlap 10a or 10b (not just spanning)
cryptic_all_struct <- all_reads %>%
  filter(includes_10a | includes_10b) %>%
  rowwise() %>%
  mutate(
    # Check for splice junction between 10a and 10b (discrete cassette)
    has_10a_10b_junction = any(
      abs(junctions$donor - EXON10A_END) <= JUNC_TOL &
      abs(junctions$acceptor - EXON10B_START) <= JUNC_TOL
    ),
    # Check for entry junction from exon 9 into 10a
    has_e9_to_10a = any(
      abs(junctions$donor - EXON9_END) <= JUNC_TOL &
      abs(junctions$acceptor - EXON10A_START) <= JUNC_TOL
    )
  ) %>%
  ungroup() %>%
  mutate(
    structure = case_when(
      # Structure 2: discrete 10a + 10b cassettes (has splice between them)
      includes_10a & includes_10b & has_10a_10b_junction
        ~ "S2: 9 \u2192 10a \u2192 10b \u2192 11",
      # Structure 3: continuous block through 10a and 10b (no splice between)
      includes_10a & includes_10b & !has_10a_10b_junction
        ~ "S3: 9 \u2192 [10a...10b] \u2192 11",
      # Structure 1: 10a only (with or without spanning)
      includes_10a & !includes_10b
        ~ "S1: 9 \u2192 10a \u2192 11",
      # Structure 4: 10b only (local reads)
      !includes_10a & includes_10b
        ~ "S4: local 10b",
      TRUE ~ "other"
    )
  )

# Count by structure, treatment, and TSS
struct_summary <- cryptic_all_struct %>%
  dplyr::count(structure, treatment) %>%
  pivot_wider(names_from = treatment, values_from = n, values_fill = 0) %>%
  mutate(Total = DMSO + Smg1i) %>%
  arrange(structure)

struct_by_tss <- cryptic_all_struct %>%
  dplyr::count(structure, treatment, tss_category) %>%
  pivot_wider(names_from = tss_category, values_from = n, values_fill = 0)
```

```{r structure-table}
struct_summary %>%
  kable(caption = "Read counts by isoform structure and treatment",
        col.names = c("Structure", "DMSO", "Smg1i", "Total"))
```

```{r structure-detail-table}
struct_by_tss %>%
  arrange(structure, treatment) %>%
  kable(caption = "Structure counts broken down by TSS category",
        col.names = c("Structure", "Treatment", "Full-length", "Internal", "Local"))
```

```{r fig7-structure-barplot, fig.height=5, fig.cap="Figure 7: Read counts by cryptic exon isoform structure. Structures 2 and 4 are enriched under Smg1i."}
struct_plot_data <- cryptic_all_struct %>%
  dplyr::count(structure, treatment)

ggplot(struct_plot_data, aes(x = structure, y = n, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.7),
    vjust = -0.3, size = 3.5
  ) +
  scale_fill_manual(values = treatment_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Cryptic exon isoform structures",
    subtitle = "Read counts across all cell types",
    x = NULL, y = "Number of reads", fill = "Treatment"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(size = 10))
```

**Structure 1 (9 → 10a → 11)** is the most common cryptic pattern. Reads show clean
canonical splice sites (8,807,060 → 8,809,448 and 8,809,561 → 8,812,979) and are found on
both internal-TSS and full-length transcripts. This matches the exon structure of MTCL1-206.

**Structure 2 (9 → 10a → 10b → 11)** treats 10a and 10b as discrete cassette exons
separated by a ~410 bp intron (8,809,561 → 8,809,971). This architecture appears
Smg1i-enriched, though absolute counts are small, and is not represented in any GENCODE
isoform.

**Structure 3 (9 → [10a...10b continuous] → 11)** has a single uninterrupted aligned block
from the 10a start through and past 10b (sometimes to 8,810,496), then splices to exon 11.
It partially resembles MTCL1-208's terminal exon but differs in continuing to exon 11.

**Structure 4 (local 10b)** consists entirely of local-TSS reads starting within the intron
between 10a and 10b (~8,809,593--8,810,180). These are markedly Smg1i-enriched and may
represent stabilized fragments of NMD-targeted transcripts.

## PTC sequence analysis

To determine whether the cryptic exon structures introduce premature termination codons
(PTCs), we extract the genomic DNA sequence for each exon, construct the spliced cDNA
insert for each structure, and translate in the reading frame established by exon 9.

The GTF CDS frame annotation confirms **frame 0** for exon 9 (8,806,893--8,807,060) across
all canonical coding isoforms (MTCL1-201, -216, -217, -204), meaning the reading frame
aligns to codon boundaries at the exon 9 splice donor. Both exon 9 and exon 11 CDS entries
carry frame 0, so canonical splicing (exon 9 → exon 11) maintains the reading frame.

```{r ptc-sequence-extraction}
# Extract genomic sequences for each region of interest
genome <- BSgenome.Hsapiens.UCSC.hg38

seq_exon9  <- getSeq(genome, "chr18", EXON9_START, EXON9_END)
seq_10a    <- getSeq(genome, "chr18", EXON10A_START, EXON10A_END)
seq_intron <- getSeq(genome, "chr18", EXON10A_END + 1, EXON10B_START - 1)  # 8809562–8809970
seq_10b    <- getSeq(genome, "chr18", EXON10B_START, EXON10B_END)
seq_exon11 <- getSeq(genome, "chr18", EXON11_START, EXON11_END)

# Report insert lengths
insert_lengths <- tibble(
  Region = c("Exon 10a", "Intron 10a-10b", "Exon 10b"),
  Length_bp = c(length(seq_10a), length(seq_intron), length(seq_10b)),
  `Mod 3` = Length_bp %% 3,
  `Frame effect` = ifelse(`Mod 3` == 0, "In-frame", paste0("Frameshift (+", `Mod 3`, ")"))
)

# Construct spliced cDNA inserts for each structure
# We translate just the insert portion in the frame established at exon 9 donor
# Exon 9 ends at a codon boundary (frame 0), so the first base of any insert is frame position 1

# Canonical skip: no insert (control — exon 9 end joins exon 11 start directly)
insert_canonical <- DNAString("")

# S1: exon 10a only (114 bp, in-frame)
insert_s1 <- DNAString(as.character(seq_10a))

# S2: exon 10a + exon 10b (114 + 228 = 342 bp, in-frame)
insert_s2 <- DNAString(paste0(as.character(seq_10a), as.character(seq_10b)))

# S3: exon 10a + intron + exon 10b (114 + 409 + 228 = 751 bp, frameshift)
insert_s3 <- DNAString(paste0(as.character(seq_10a), as.character(seq_intron),
                               as.character(seq_10b)))

# Function to translate an insert and find first PTC
find_ptc <- function(insert_seq, structure_name) {
  insert_len <- length(insert_seq)

  if (insert_len == 0) {
    return(tibble(
      Structure = structure_name,
      Insert_bp = 0,
      Frame_status = "In-frame (no insert)",
      PTC_found = FALSE,
      PTC_codon_pos = NA_integer_,
      PTC_nt_pos = NA_integer_,
      AA_context = NA_character_,
      Translation = NA_character_
    ))
  }

  # Pad to complete codons for translation
  pad_len <- (3 - (insert_len %% 3)) %% 3
  if (pad_len > 0) {
    # Borrow bases from exon 11 start to complete last codon
    pad_bases <- subseq(seq_exon11, 1, pad_len)
    translate_seq <- DNAString(paste0(as.character(insert_seq), as.character(pad_bases)))
  } else {
    translate_seq <- insert_seq
  }

  aa_seq <- translate(translate_seq, no.init.codon = TRUE)
  aa_str <- as.character(aa_seq)

  # Find first stop codon
  stop_pos <- regexpr("\\*", aa_str)
  ptc_found <- stop_pos > 0

  if (ptc_found) {
    ptc_codon <- as.integer(stop_pos)
    ptc_nt <- (ptc_codon - 1) * 3 + 1  # nucleotide position within insert

    # Extract amino acid context (±5 AA around PTC)
    ctx_start <- max(1, ptc_codon - 5)
    ctx_end <- min(nchar(aa_str), ptc_codon + 5)
    context <- substr(aa_str, ctx_start, ctx_end)
  } else {
    ptc_codon <- NA_integer_
    ptc_nt <- NA_integer_
    context <- NA_character_
  }

  frame_status <- if (insert_len %% 3 == 0) "In-frame" else
    paste0("Frameshift (+", insert_len %% 3, ")")

  tibble(
    Structure = structure_name,
    Insert_bp = insert_len,
    Frame_status = frame_status,
    PTC_found = ptc_found,
    PTC_codon_pos = ptc_codon,
    PTC_nt_pos = ptc_nt,
    AA_context = context,
    Translation = aa_str
  )
}

# Analyze each structure
ptc_results <- bind_rows(
  find_ptc(insert_canonical, "Canonical skip (9→11)"),
  find_ptc(insert_s1, "S1: 9→10a→11"),
  find_ptc(insert_s2, "S2: 9→10a→10b→11"),
  find_ptc(insert_s3, "S3: 9→[10a...10b]→11")
)
```

```{r ptc-summary-table}
ptc_display <- ptc_results %>%
  mutate(
    `PTC found?` = ifelse(PTC_found, "Yes", "No"),
    `PTC codon position` = ifelse(is.na(PTC_codon_pos), "—",
                                   as.character(PTC_codon_pos)),
    `AA context (±5)` = ifelse(is.na(AA_context), "—", AA_context)
  ) %>%
  select(Structure, `Insert length (bp)` = Insert_bp, `Frame status` = Frame_status,
         `PTC found?`, `PTC codon position`, `AA context (±5)`)

ptc_display %>%
  kable(caption = "PTC analysis of cryptic exon structures. Codon position is relative to the start of the insert sequence, in the reading frame established by exon 9.")
```

```{r ptc-sequence-viz, fig.height=4, fig.cap="Figure 8: PTC map of cryptic exon inserts. Each bar represents the translated insert for one structure. Green = translated amino acids before PTC; red triangle = PTC position; gray = sequence after PTC. Bar length proportional to insert size."}
# Build visualization data for structures with inserts
ptc_viz <- ptc_results %>%
  filter(Insert_bp > 0) %>%
  mutate(
    total_codons = ceiling(Insert_bp / 3),
    pre_ptc = ifelse(PTC_found, PTC_codon_pos - 1, total_codons),
    y_pos = rev(row_number())
  )

# Create rectangles for translated (green) and post-PTC (gray) regions
viz_rects <- ptc_viz %>%
  rowwise() %>%
  mutate(
    rects = list(bind_rows(
      # Pre-PTC (translated)
      tibble(xmin = 0, xmax = pre_ptc, region = "Translated", y = y_pos),
      # Post-PTC (if PTC found)
      if (PTC_found)
        tibble(xmin = PTC_codon_pos, xmax = total_codons, region = "After PTC", y = y_pos)
      else
        tibble(xmin = numeric(0), xmax = numeric(0), region = character(0), y = numeric(0))
    ))
  ) %>%
  unnest(rects)

# PTC markers
ptc_markers <- ptc_viz %>%
  filter(PTC_found)

region_colors <- c("Translated" = "#27AE60", "After PTC" = "#BDC3C7")

ggplot() +
  geom_rect(
    data = viz_rects,
    aes(xmin = xmin, xmax = xmax, ymin = y - 0.35, ymax = y + 0.35, fill = region),
    color = "grey40", linewidth = 0.3
  ) +
  geom_point(
    data = ptc_markers,
    aes(x = PTC_codon_pos, y = y_pos),
    shape = 25, size = 4, fill = "#E74C3C", color = "#E74C3C"
  ) +
  geom_text(
    data = ptc_markers,
    aes(x = PTC_codon_pos, y = y_pos + 0.55,
        label = paste0("PTC @ codon ", PTC_codon_pos)),
    size = 3, color = "#E74C3C", fontface = "bold"
  ) +
  scale_fill_manual(values = region_colors, name = "Region") +
  scale_y_continuous(
    breaks = ptc_viz$y_pos,
    labels = ptc_viz$Structure
  ) +
  labs(
    title = "PTC map of cryptic exon inserts",
    subtitle = "Reading frame established by exon 9 (CDS frame 0)",
    x = "Codon position within insert", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10)
  )
```

## PTC enrichment in NMD-inhibited cells

Having determined the PTC status of each structure from the genomic sequence, we can now
classify every spanning read as PTC-containing or non-PTC and test whether PTC reads are
enriched under Smg1i treatment. This is conceptually distinct from the overall cryptic exon
inclusion test (Section 4.2): that test asks whether cryptic exons are included more often
under Smg1i, while this test asks whether **PTC-containing** structures are specifically
enriched --- a more direct test of NMD.

```{r ptc-read-classification}
# Determine PTC status for each structure from sequence analysis
s1_has_ptc <- ptc_results %>% filter(Structure == "S1: 9→10a→11") %>% pull(PTC_found)
s2_has_ptc <- ptc_results %>% filter(Structure == "S2: 9→10a→10b→11") %>% pull(PTC_found)
# S3 always has PTC (frameshift guarantees it)
s3_has_ptc <- TRUE

# Classify spanning reads by PTC status
# Start from spanning tibble (reads that span exon 9 to exon 11)
spanning_ptc <- spanning %>%
  rowwise() %>%
  mutate(
    # Check for splice junction between 10a and 10b (discrete cassette)
    has_10a_10b_junction = any(
      abs(junctions$donor - EXON10A_END) <= JUNC_TOL &
      abs(junctions$acceptor - EXON10B_START) <= JUNC_TOL
    ),
    # Check for entry from exon 9 into 10a
    has_e9_to_10a = any(
      abs(junctions$donor - EXON9_END) <= JUNC_TOL &
      abs(junctions$acceptor - EXON10A_START) <= JUNC_TOL
    )
  ) %>%
  ungroup() %>%
  mutate(
    read_structure = case_when(
      includes_10a & includes_10b & has_10a_10b_junction ~ "S2",
      includes_10a & includes_10b & !has_10a_10b_junction ~ "S3",
      includes_10a & !includes_10b ~ "S1",
      !includes_10a & includes_10b ~ "S4",
      TRUE ~ "canonical"
    ),
    ptc_status = case_when(
      read_structure == "canonical" ~ "non-PTC",
      read_structure == "S1" & s1_has_ptc ~ "PTC",
      read_structure == "S1" & !s1_has_ptc ~ "non-PTC",
      read_structure == "S2" & s2_has_ptc ~ "PTC",
      read_structure == "S2" & !s2_has_ptc ~ "non-PTC",
      read_structure == "S3" ~ "PTC",  # frameshift → guaranteed PTC
      read_structure == "S4" ~ NA_character_  # exclude: no canonical frame context
    )
  )

# Summary of PTC classification
ptc_class_summary <- spanning_ptc %>%
  filter(!is.na(ptc_status)) %>%
  dplyr::count(read_structure, ptc_status) %>%
  arrange(read_structure)

ptc_class_summary %>%
  kable(caption = "PTC classification of spanning reads by structure type. S4 reads excluded (no canonical frame context).",
        col.names = c("Structure", "PTC status", "N reads"))
```

```{r ptc-enrichment-global}
# Global 2×2 contingency table: PTC vs non-PTC × Smg1i vs DMSO
ptc_for_test <- spanning_ptc %>%
  filter(!is.na(ptc_status))

global_table <- ptc_for_test %>%
  dplyr::count(treatment, ptc_status) %>%
  pivot_wider(names_from = ptc_status, values_from = n, values_fill = 0)

# Build matrix for Fisher's test: rows = treatment, cols = PTC status
mat_global <- matrix(
  c(global_table %>% filter(treatment == "Smg1i") %>% pull(PTC),
    global_table %>% filter(treatment == "Smg1i") %>% pull(`non-PTC`),
    global_table %>% filter(treatment == "DMSO") %>% pull(PTC),
    global_table %>% filter(treatment == "DMSO") %>% pull(`non-PTC`)),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("Smg1i", "DMSO"), c("PTC", "non-PTC"))
)

fisher_global <- fisher.test(mat_global)

cat("Global PTC enrichment test (Fisher's exact):\n")
cat(sprintf("  Contingency table:\n"))
print(mat_global)
cat(sprintf("\n  Odds ratio: %.2f (95%% CI: %.2f–%.2f)\n",
            fisher_global$estimate,
            fisher_global$conf.int[1],
            fisher_global$conf.int[2]))
cat(sprintf("  P-value: %.4g\n", fisher_global$p.value))
cat(sprintf("  Interpretation: %s\n",
            ifelse(fisher_global$estimate > 1 & fisher_global$p.value < 0.05,
                   "PTC-containing reads are significantly enriched under Smg1i",
                   ifelse(fisher_global$estimate > 1,
                          "PTC-containing reads trend toward enrichment under Smg1i (not significant)",
                          "No enrichment of PTC reads under Smg1i"))))
```

```{r ptc-enrichment-celltype}
# Per-cell-type Fisher's tests
celltype_ptc_counts <- ptc_for_test %>%
  dplyr::count(cell_type, treatment, ptc_status) %>%
  complete(cell_type, treatment, ptc_status, fill = list(n = 0))

celltype_fisher <- celltype_ptc_counts %>%
  pivot_wider(names_from = c(treatment, ptc_status), values_from = n, values_fill = 0) %>%
  rowwise() %>%
  mutate(
    n_spanning = sum(c_across(starts_with("Smg1i") | starts_with("DMSO"))),
    mat = list(matrix(
      c(`Smg1i_PTC`, `Smg1i_non-PTC`, `DMSO_PTC`, `DMSO_non-PTC`),
      nrow = 2, byrow = TRUE
    )),
    fisher = list(fisher.test(mat)),
    OR = fisher$estimate,
    ci_low = fisher$conf.int[1],
    ci_high = fisher$conf.int[2],
    p_value = fisher$p.value,
    min_arm = min(Smg1i_PTC + `Smg1i_non-PTC`, DMSO_PTC + `DMSO_non-PTC`),
    underpowered = min_arm < 10
  ) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    sig = case_when(
      underpowered ~ "underpowered",
      p_adj < 0.05 ~ "significant",
      TRUE ~ "ns"
    )
  ) %>%
  select(cell_type, Smg1i_PTC, `Smg1i_non-PTC`, DMSO_PTC, `DMSO_non-PTC`,
         OR, ci_low, ci_high, p_value, p_adj, underpowered, sig)

celltype_fisher %>%
  mutate(
    OR_display = sprintf("%.2f [%.2f–%.2f]", OR, ci_low, ci_high),
    p_display = sprintf("%.4g", p_adj),
    flag = ifelse(underpowered, "*", "")
  ) %>%
  select(cell_type, Smg1i_PTC, `Smg1i_non-PTC`, DMSO_PTC, `DMSO_non-PTC`,
         `OR [95% CI]` = OR_display, `FDR p` = p_display, Flag = flag) %>%
  kable(caption = "Per-cell-type PTC enrichment (Fisher's exact test, BH-corrected). * = underpowered (<10 spanning reads per arm).")
```

```{r ptc-forest-plot, fig.height=5, fig.cap="Figure 9: Forest plot of PTC enrichment across cell types. Log odds ratio (x-axis) with 95% CI. Vertical dashed line at OR=1 (no enrichment). Pooled estimate shown at bottom. Gray points indicate underpowered cell types (<10 reads per arm)."}
# Combine global and per-cell-type results for forest plot
forest_data <- bind_rows(
  celltype_fisher %>%
    select(cell_type, OR, ci_low, ci_high, underpowered) %>%
    dplyr::rename(label = cell_type),
  tibble(
    label = "Pooled",
    OR = fisher_global$estimate,
    ci_low = fisher_global$conf.int[1],
    ci_high = fisher_global$conf.int[2],
    underpowered = FALSE
  )
) %>%
  mutate(
    log_OR = log(OR),
    log_ci_low = log(ci_low),
    log_ci_high = log(ci_high),
    # Cap infinite CIs for plotting
    log_ci_low = pmax(log_ci_low, -5),
    log_ci_high = pmin(log_ci_high, 5),
    y_pos = rev(row_number()),
    is_pooled = label == "Pooled",
    point_color = case_when(
      is_pooled ~ "#000000",
      underpowered ~ "#95A5A6",
      TRUE ~ "#2C3E50"
    )
  )

ggplot(forest_data, aes(x = log_OR, y = y_pos)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(
    aes(xmin = log_ci_low, xmax = log_ci_high),
    height = 0.2, color = forest_data$point_color
  ) +
  geom_point(
    aes(shape = is_pooled),
    size = 3, fill = forest_data$point_color, color = forest_data$point_color
  ) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 18), guide = "none") +
  scale_y_continuous(
    breaks = forest_data$y_pos,
    labels = forest_data$label
  ) +
  labs(
    title = "PTC enrichment under Smg1i",
    subtitle = "Log odds ratio with 95% CI (Fisher's exact test)",
    x = "log(Odds Ratio)", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(
      face = ifelse(forest_data$is_pooled[order(-forest_data$y_pos)], "bold", "plain"),
      size = 11
    )
  )
```

## NMD mechanism

The Smg1i enrichment of cryptic exon inclusion --- particularly the discrete 10a + 10b
cassette (Structure 2) and the local 10b reads (Structure 4) --- is consistent with
**NMD-mediated degradation** of these transcripts under normal conditions.

```{r nmd-mechanism-summary, results='asis'}
# Generate definitive PTC summary based on sequence analysis results
s1_ptc <- ptc_results %>% filter(Structure == "S1: 9→10a→11") %>% pull(PTC_found)
s2_ptc <- ptc_results %>% filter(Structure == "S2: 9→10a→10b→11") %>% pull(PTC_found)
s1_pos <- ptc_results %>% filter(Structure == "S1: 9→10a→11") %>% pull(PTC_codon_pos)
s2_pos <- ptc_results %>% filter(Structure == "S2: 9→10a→10b→11") %>% pull(PTC_codon_pos)
s3_pos <- ptc_results %>% filter(Structure == "S3: 9→[10a...10b]→11") %>% pull(PTC_codon_pos)

# Count structures with confirmed PTCs
n_ptc_structures <- sum(c(s1_ptc, s2_ptc, TRUE))  # S3 always has PTC

cat("Sequence analysis (Section 5.3) **confirms PTCs** in the cryptic exon inserts:\n\n")

if (s1_ptc) {
  cat(sprintf("- **S1 (9→10a→11)**: In-frame stop codon at **codon %d** within the 10a insert (114 bp, in-frame)\n", s1_pos))
} else {
  cat("- **S1 (9→10a→11)**: No in-frame stop codon in the 10a insert (114 bp) — this structure does **not** trigger NMD via a PTC in the cryptic exon\n")
}

if (s2_ptc) {
  cat(sprintf("- **S2 (9→10a→10b→11)**: In-frame stop codon at **codon %d** within the 10a+10b insert (342 bp, in-frame)\n", s2_pos))
} else {
  cat("- **S2 (9→10a→10b→11)**: No in-frame stop codon in the 10a+10b insert (342 bp) — this structure does **not** trigger NMD via a PTC in the cryptic exon\n")
}

cat(sprintf("- **S3 (9→[10a...10b]→11)**: Retained intron (410 bp, 410%%3 = 2) causes a **frameshift**, introducing a PTC at **codon %d** within the insert\n\n", s3_pos))

cat(sprintf("In total, **%d of 3** spliced cryptic exon structures contain a PTC. ", n_ptc_structures))
cat("All PTCs are located thousands of nucleotides upstream of the last exon–exon junction, ")
cat("satisfying the **50-nt rule** for NMD substrate recognition.\n\n")

cat("**Note on MTCL1-208**: GENCODE's NMD-annotated isoform (MTCL1-208) uses a different CDS start ")
cat("(8,783,587 vs. the canonical 8,705,661) and its PTC at genomic position 8,796,309 is upstream ")
cat("of the cryptic exon region. The novel structures S1–S4 use the canonical CDS start, so their ")
cat("NMD-triggering mechanism is independent of MTCL1-208's.\n")
```

## Cell-type specificity

The cryptic exon inclusion effect is strongest in **DD** and **DD_ALI**, suggesting that
NMD-mediated regulation of these exons is most active in differentiated airway cell types.
MV cells show the opposite trend (decreased inclusion under Smg1i), indicating cell-type-specific
regulation of this splicing event.

```{r celltype-ptc-integration, results='asis'}
# Report which cell types show significant PTC enrichment
sig_cts <- celltype_fisher %>% filter(sig == "significant") %>% pull(cell_type)
underpowered_cts <- celltype_fisher %>% filter(underpowered) %>% pull(cell_type)

if (length(sig_cts) > 0) {
  cat(sprintf("Per-cell-type PTC enrichment analysis (Section 5.4) confirms significant enrichment in **%s** (BH-adjusted p < 0.05). ",
              paste(sig_cts, collapse = ", ")))
} else {
  cat("Per-cell-type PTC enrichment tests did not reach significance individually after FDR correction, ")
  cat("likely due to limited per-cell-type read counts. ")
}

if (length(underpowered_cts) > 0) {
  cat(sprintf("Cell types **%s** had fewer than 10 spanning reads per treatment arm and are underpowered for this comparison. ",
              paste(underpowered_cts, collapse = ", ")))
}

cat("The pooled analysis across all cell types provides the most robust estimate of PTC enrichment (see Figure 9).\n")
```

## Limitations

- **Merged BAMs**: Each BAM represents merged data across donors per cell type/treatment,
  so donor-level variability cannot be assessed.
- **Small read counts**: Total reads across all BAMs are ~2,000, limiting statistical power
  for rare events and making per-cell-type comparisons imprecise.
- **Long-read quantification**: PacBio IsoSeq capture efficiency may vary across transcript
  lengths and structures, potentially biasing inclusion rate estimates.
- **Local reads**: A substantial fraction of cryptic-exon-including reads are locally-initiating
  transcripts (Structure 4). Whether these represent bona fide NMD substrates or capture/
  fragmentation artifacts requires further investigation.
- **Transcript model mismatch**: The cryptic exon structures observed in the reads do not
  correspond to existing GENCODE or PacBio-called isoform models. Quantification pipelines
  that rely on transcript-level assignment will miss or misassign these reads.
- **S4 excluded from PTC enrichment**: Structure 4 (local 10b) reads lack canonical frame
  context (no exon 9 splice donor) and are excluded from the PTC enrichment analysis. Their
  NMD relevance is inferred from Smg1i enrichment rather than confirmed PTC status.
- **Junction tolerance and frame assumption**: PTC classification assumes reads matching
  canonical splice sites (±5 bp tolerance) preserve the exon 9 reading frame. Reads with
  non-canonical splice sites could have different frame contexts.
- **Conservative enrichment test**: The PTC enrichment analysis compares PTC vs non-PTC reads
  among spanning reads only; it does not capture non-spanning PTC transcripts that may be
  fully degraded under DMSO.
- **Platform comparison caveats**: rMATS can only test exon 10a; 10b and multi-exon cryptic
  structures are annotation-dependent and not detected. Long-read and short-read denominators
  differ (read-spanning vs junction-count-based).

# Session Info

```{r session-info}
sessionInfo()
```
